Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-09 conclusions.tex
Converted to AASTeX
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@997 ccdd05d5-56c1-461a-bf98-85f6876523b4

\section{Conclusions}
\label{SECV}\label{sec:conclusions}

We have demonstrated a computationally feasible procedure for the rapid and
even preemptive detection of \GW{}s emitted during the coalescence
of neutron stars and stellar-mass black holes. Our method is as fast as
standard \fft{} convolutions but allows for zero latency, sample-in-sample-out
operation.  Our search targets are expected to produce prompt electromagnetic
signals and may be the progenitors of some short hard gamma-ray bursts.  Rapid
alerts to the broader astronomical community will improve the chances of
detecting an electromagnetic counterpart in bands from gamma-rays down to
radio.  We anticipate requiring $\sim$100 modern multi-core computers to
analyze a four-detector network of \GW{} data for binary neutron
stars and stellar-mass black holes.  This is well within the current computing
capabilities of the \LIGO{} Data Grid~\cite{LDG}. In the future, this could be
improved upon further by conditionally reconstructing the \SNR{} time-series
only during times when a composite detection statistic crosses a
threshold~\cite{svd-compdetstat}.

The algorithm we described has no intrinsic latency.  However, there are
fundamental and practical latencies associated with the analysis and detection
procedure. For example, the \LIGO{} detectors, data acquisition is synchronized
to a 1/(16~Hz) cadence, introducing an up-front latency~\cite{Bork2001}. Data
aggregation from the observatories will travel over various networks, each
capable of high bandwidth but perhaps only modest latency.  This could amount to
a similar latency of $\sim$100 ms.  Lastly, unless a real-time infrastructure
is adopted post data acquisition, it is likely that there will be an inherent
latency introduced by such infrastructure.  We have shown a prototype
implementation of \lloid{} using the open source signal processing software
\gstreamer\ and \gstlal. Significant work would have to be done in order to
improve upon the latency capability of our implementation, for example, more
tightly integrating analysis and data acquisition. This should be considered
for third-generation detector design.

Although we have demonstrated a feasible method for advance detection we have
not explored the accuracy of sky localization that is possible before merger.
Ref.~\cite{Fairhurst2009} discusses some of the theoretical prospects for early
sky localization.  Our future work will explore the prospects of early-warning
detection with realistic simulations of binary mergers using the infrastructure
and techniques described here.

>>>>

\section{Conclusions}
\label{sec:conclusions}

We have demonstrated a computationally feasible procedure for the rapid and
even preemptive detection of \GW{}s emitted during the coalescence
of neutron stars and stellar-mass black holes. Our method is as fast as
standard \fft{} convolutions but allows for zero latency, sample-in-sample-out
operation.  Our search targets are expected to produce prompt electromagnetic
signals and may be the progenitors of some short hard gamma-ray bursts.  Rapid
alerts to the broader astronomical community will improve the chances of
detecting an electromagnetic counterpart in bands from gamma-rays down to
radio.  We anticipate requiring $\sim$100 modern multi-core computers to
analyze a four-detector network of \GW{} data for binary neutron
stars and stellar-mass black holes.  This is well within the current computing
capabilities of the \LIGO{} Data Grid~\citep{LDG}. In the future, this could be
improved upon further by conditionally reconstructing the \SNR{} time-series
only during times when a composite detection statistic crosses a
threshold~\citep{svd-compdetstat}.

The algorithm we described has no intrinsic latency.  However, there are
fundamental and practical latencies associated with the analysis and detection
procedure. For example, the \LIGO{} detectors, data acquisition is synchronized
to a 1/(16~Hz) cadence, introducing an up-front latency~\citep{Bork2001}. Data
aggregation from the observatories will travel over various networks, each
capable of high bandwidth but perhaps only modest latency.  This could amount to
a similar latency of $\sim$100 ms.  Lastly, unless a real-time infrastructure
is adopted post data acquisition, it is likely that there will be an inherent
latency introduced by such infrastructure.  We have shown a prototype
implementation of \lloid{} using the open source signal processing software
\gstreamer\ and \gstlal. Significant work would have to be done in order to
improve upon the latency capability of our implementation, for example, more
tightly integrating analysis and data acquisition. This should be considered
for third-generation detector design.

Although we have demonstrated a feasible method for advance detection we have
not explored the accuracy of sky localization that is possible before merger.
Ref.~\citep{Fairhurst2009} discusses some of the theoretical prospects for early
sky localization.  Our future work will explore the prospects of early-warning
detection with realistic simulations of binary mergers using the infrastructure
and techniques described here.

### 2011-06-09 implementation.tex
Converted to AASTeX
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@997 ccdd05d5-56c1-461a-bf98-85f6876523b4

The planning stage begins with choosing templates that cover the space of
source parameters with a hexagonal grid~\cite{PhysRevD.76.102004} in order to
satisfy a minimum match criterion.  This assures a prescribed maximum loss in
\SNR\ for signals that fall between the chosen templates.  Typically the
minimum match is 0.97 corresponding to a maximum mismatch of 0.03.  Next, the
templates are subdivided into groups of neighbors called \emph{sub-banks} that
are appropriately sized so that each bank can be efficiently handled by a
single computer.  The neighbors are chosen to have comparable chirp mass, which
produces sub-banks with similar time-frequency evolution.  Dividing the source
parameter space into smaller sub-banks reduces the computational cost of the
\SVD{} and is the approach considered
in~\cite{Cannon:2010p10398}.  We choose time-slice boundaries as in
equation~\eqref{eq:time-sliced-templates} such that all of the templates within
a sub-bank are sub-critically sampled at progressively lower sample rates.
Next, the templates within the sub-bank are realized as \fir\ filter
coefficients.  For each time slice, the templates are down-sampled to the
appropriate sample rate.  Finally, the \SVD\ is applied to each time slice in
the sub-bank in order to produce a set of orthogonal \fir\ filters and a
reconstruction matrix that maps them back to the original templates as
described in equation~\eqref{eq:svddecomp}.  The down-sampled orthogonal \fir\
filter coefficients, the reconstruction matrix, and the time-slice boundaries
are all saved to disk.

>>>>

The planning stage begins with choosing templates that cover the space of
source parameters with a hexagonal grid~\citep{PhysRevD.76.102004} in order to
satisfy a minimum match criterion.  This assures a prescribed maximum loss in
\SNR\ for signals that fall between the chosen templates.  Typically the
minimum match is 0.97 corresponding to a maximum mismatch of 0.03.  Next, the
templates are subdivided into groups of neighbors called \emph{sub-banks} that
are appropriately sized so that each bank can be efficiently handled by a
single computer.  The neighbors are chosen to have comparable chirp mass, which
produces sub-banks with similar time-frequency evolution.  Dividing the source
parameter space into smaller sub-banks reduces the computational cost of the
\SVD{} and is the approach considered
in~\citep{Cannon:2010p10398}.  We choose time-slice boundaries as in
equation~\eqref{eq:time-sliced-templates} such that all of the templates within
a sub-bank are sub-critically sampled at progressively lower sample rates.
Next, the templates within the sub-bank are realized as \fir\ filter
coefficients.  For each time slice, the templates are down-sampled to the
appropriate sample rate.  Finally, the \SVD\ is applied to each time slice in
the sub-bank in order to produce a set of orthogonal \fir\ filters and a
reconstruction matrix that maps them back to the original templates as
described in equation~\eqref{eq:svddecomp}.  The down-sampled orthogonal \fir\
filter coefficients, the reconstruction matrix, and the time-slice boundaries
are all saved to disk.

### 2011-06-09 inspiral_svd.tex
added funding ack
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@957 ccdd05d5-56c1-461a-bf98-85f6876523b4
Updates
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@971 ccdd05d5-56c1-461a-bf98-85f6876523b4
Converted to AASTeX
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@997 ccdd05d5-56c1-461a-bf98-85f6876523b4

\documentclass{iopart}
\input{packages}
\input{macros}
\renewcommand{\arraystretch}{1.2}

@@

\title[Early-warning detection of compact binary coalescence]{Technique for \earlywarning\ detection of compact binary
coalescence and its implications for multi-messenger astronomy}

\date{\today}

\author{Kipp Cannon$^{1}$, Romain Cariou$^{2}$, Adrian Chapman$^3$, Mireia
Crisp\'{i}n-Ortuzar$^4$, Nickolas Fotopoulos$^3$, Melissa Frei $^5$, Chad
Hanna$^{6}$, Erin Kara$^{7}$, Drew Keppel$^{8,9}$, Laura Liao$^{10}$, Stephen
Privitera $^3$, Antony C Searle$^3$, Leo Singer$^3$ and Alan J  Weinstein$^3$}

\address{$^1$ Canadian Institute for Theoretical Astrophysics, Toronto, ON,
Canada}
\address{$^2$  D\'{e}partement de physique, \'{E}cole Normale Sup\'{e}rieure de Cachan, Cachan, France}
\address{$^3$ LIGO Laboratory - California Institute of Technology, Pasadena,
CA, USA} 
\address{$^4$ Facultat de F\'{i}sica, Universitat de Val\`{e}ncia, Burjassot,
Spain} 
\address{$^5$ The University of Texas at Austin, Austin, TX, USA}
\address{$^6$ Perimeter Institute for Theoretical Physics, Waterloo, ON,
Canada} 
\address{$^7$ Department of Physics and Astronomy, Barnard College, Columbia
University, New York, NY, USA} 
\address{$^8$ Albert-Einstein-Institut, Max-Planck-Institut f\"{u}r
Gravitationphysik, Hannover, Germany} 
\address{$^9$ Leibniz Universit\"{a}t Hannover, Hannover, Germany}
\address{$^{10}$ Ryerson University, Toronto, ON, Canada}

@@

\pacs{04.30.-w, 07.05.Kf}

%\preprint{}
\input{introduction}

\input{method}

\input{implementation}

\input{results}

\input{conclusions}

\ack{\LIGO\ was constructed by the California Institute of Technology and
Massachusetts Institute of Technology with funding from the National Science
Foundation and operates under cooperative agreement
\textsc{phy}-\oldstylenums{0107417}.  CH would like to thank Ilya Mandel for
many discussions about rate estimates and the prospects of early detection.  CH
would also like to thank Patrick Brady for countless fruitful conversations
about low latency analysis methods.  NF would like to thank Alessandra Corsi
for illuminating discussions on astronomical motivations.  This research is
supported by the National Science Foundation through a Graduate Research
Fellowship to LS.

This paper has \LIGO\ Document
Number \LIGO-\textsc{p}\oldstylenums{0900004}.}

%\appendix
%\input{appendix}

\section*{References}

\bibliographystyle{iopart-num}
\bibliography{references}

>>>>

\documentclass[preprint2]{aastex}
\input{packages}
\input{macros}

@@

\title{Technique for \earlywarning\ detection of compact binary coalescence
and its implications for multi-messenger astronomy}

\shorttitle{Early-warning of compact binary coalescence}

\author{
	Kipp Cannon\altaffilmark{1},
	Romain Cariou\altaffilmark{2},
	Adrian Chapman\altaffilmark{3},
	Mireia Crisp\'{i}n-Ortuzar\altaffilmark{4},
	Nickolas Fotopoulos\altaffilmark{3},
	Melissa Frei\altaffilmark{5},
	Chad Hanna\altaffilmark{6},
	Erin Kara\altaffilmark{7},
	Drew Keppel\altaffilmark{8,9},
	Laura Liao\altaffilmark{10},
	Stephen Privitera\altaffilmark{3},
	Antony Searle\altaffilmark{3},
	Leo Singer\altaffilmark{3}, and
	Alan Weinstein\altaffilmark{3}
}

\altaffiltext{1}{Canadian Institute for Theoretical Astrophysics, Toronto, ON, Canada}
\altaffiltext{2}{D\'{e}partement de physique, \'{E}cole Normale Sup\'{e}rieure de Cachan, Cachan, France}
\altaffiltext{3}{LIGO Laboratory - California Institute of Technology, Pasadena, CA, USA} 
\altaffiltext{4}{Facultat de F\'{i}sica, Universitat de Val\`{e}ncia, Burjassot, Spain} 
\altaffiltext{5}{University of Texas at Austin, Austin, TX, USA}
\altaffiltext{6}{Perimeter Institute for Theoretical Physics, Waterloo, ON, Canada}
\altaffiltext{7}{Department of Physics and Astronomy, Barnard College, Columbia University, New York, NY, USA}
\altaffiltext{8}{Albert-Einstein-Institut, Max-Planck-Institut f\"{u}r Gravitationphysik, Hannover, Germany}
\altaffiltext{9}{Leibniz Universit\"{a}t Hannover, Hannover, Germany}
\altaffiltext{10}{Ryerson University, Toronto, ON, Canada}

\keywords{gamma ray burst: general --- gravitational waves --- methods: data analysis --- methods: numerical}

@@

\input{introduction}
\input{method}
\input{implementation}
\input{results}
\input{conclusions}

\acknowledgements

\LIGO\ was constructed by the California Institute of Technology and
Massachusetts Institute of Technology with funding from the National Science
Foundation and operates under cooperative agreement
\textsc{phy}-\oldstylenums{0107417}.  CH would like to thank Ilya Mandel for
many discussions about rate estimates and the prospects of early detection.  CH
would also like to thank Patrick Brady for countless fruitful conversations
about low latency analysis methods.  NF would like to thank Alessandra Corsi
for illuminating discussions on astronomical motivations.  DK was supported
from the Max Planck Gesellschaft. This research is supported by the National
Science Foundation through a Graduate Research Fellowship to LS and by a
fellowship from The Perimeter Institute for Theoretical Physics to CH.

This paper has \LIGO\ Document Number \LIGO-\textsc{p}\oldstylenums{0900004}.

\bibliographystyle{apj}
\bibliography{references}

### 2011-06-09 introduction.tex
Converted to AASTeX
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@997 ccdd05d5-56c1-461a-bf98-85f6876523b4

The coalescence of compact binary systems consisting of neutron stars
 (\textsc{ns}) and/or black holes (\textsc{bh}) is presently the best
understood and most actively pursued target source of gravitational
radiation for ground-based gravitational-wave (\GW{}) detectors.
As a compact binary system loses energy to \GW{}s, its
orbital separation decays, leading to a run-away inspiral with the \GW{}
amplitude and frequency increasing until the system eventually merges.  It is
thought that if a \textsc{ns} is involved, it may become tidally disrupted near
the merger and fuel a bright electromagnetic (\EM{})
counterpart~\cite{shibata:2007}. Observation of an \EM{} counterpart to a
candidate \GW{} event will vastly boost confidence in the \GW{} detection and
will often provide sufficient sky localization to determine
the source's host galaxy and therefore measure the source's redshift. The
\GW{} observation of this event will also provide a luminosity
distance measurement. With both quantities, we can
immediately measure the Hubble constant with unprecedented
accuracy~\cite{2010ApJ...725..496N}. The most efficient means of obtaining
such joint observations is through the rapid identification and dissemination
of candidate \GW{} events.

In the era of first-generation detectors, the \GW{} community initiated a
project to send alerts when potential \GW{} transients were observed in
order to trigger followup observations by optical telescopes.  The best
demonstrated latencies were on the order of 1 hour%
~\cite{HugheyGWPAW2011}, which was an important
achievement, but far too late to catch prompt \EM{}
emission. In order to extract the maximum science from Advanced \LIGO{}
\CBC{} events, we have the ambition of reporting \GW{} candidates not minutes
\emph{after} the merger, as has already been accomplished, but seconds
\emph{before}.

Though compact binary coalescence (\CBC{}) events may be responsible for several
classes of \EM{} transients, \CBC{}s are in particular believed to be a
mechanism for short gamma-ray bursts (short \GRB{}s)~\cite{Lee:2005, nakar07}.
In this scenario, prompt \EM{} emission arises as shells of
relativistically out-flowing matter collide in the inner shock.
The same inner shocks, or potentially reverse shocks, can produce a bright
accompanying optical flash~\cite{Sari99}. Prompt emission is a probe into
the extreme initial conditions of the outflow, in contrast with afterglows,
which arise in the external shock with the local medium and are relatively
insensitive to initial conditions.

Optical flashes have only been observed for a handful of long
\GRB{}s~\cite{2011CRPhy..12..255A} by telescopes with extremely rapid response
or, in the case of \textsc{grb 080319b}, by pure serendipity, where several
telescopes were observing a previous \GRB{} in the same
field~\cite{2008Natur.455..183R}. The observed optical flashes peaked
within tens of seconds and decayed quickly.
Short \GRB{}s, on the other hand, typically
fade too quickly to even catch the tails of the afterglows in any band. Rapid
\GW{} transient alerts could enable the first observation of
prompt optical flashes and the rise of afterglows from short \GRB{}s, and, for
an exceptional event, perhaps even a glimpse of a
precursor~\cite{0004-637X-723-2-1711}.

We show that by examining the signal to noise ratio (\SNR{}) stream for
threshold crossings before the \GW{} signal leaves the detection band, it is
possible to trade some \SNR{} and sky localization
accuracy for negative latency.  Figure~\ref{fig:earlywarning} shows
projected early trigger rates for \textsc{ns}--\textsc{ns} binaries in Advanced
\LIGO\ assuming event rates predicted in~\cite{Abadie:2010p10836} and anticipated
detector sensitivity described in \cite{ALIGONoise}.  The
most likely estimates indicate that $\sim$10 sources will produce an \SNR\
above 8 in the detector 10 seconds or more prior to the merger during one
year of observation, though the uncertainty spans orders of magnitude.
The gray bands give the 5 to 95\% confidence interval as described
in~\cite{Abadie:2010p10836}.
%
\begin{figure}
\begin{center}
\includegraphics{figures/snr_in_time.pdf}
\caption{\label{fig:earlywarning} Expected number of \textsc{ns}--\textsc{ns}
sources that will be detectable $t$ seconds before coalescence.  The heavy
solid line is the most likely yearly rate estimate $\dot N_{\mathrm{re}}$
during Advanced \LIGO.  The shaded region represents the confidence interval
$[\dot N_{\mathrm{low}}, \dot N_{\mathrm{high}}]$ described in
\cite{Abadie:2010p10836}.  With an advanced detector in the `zero detuning,
high power' configuration \cite{ALIGONoise} and an \SNR\ threshold of 8,
we will observe a total of $\dot N_{\mathrm{re}} = 40$ events~yr$^{-1}$, but
$\sim$10~yr$^{-1}$ will be detectable within 10~s of merger and
$\sim$1~yr$^{-1}$ within 100~s.} 
\end{center}
\end{figure}

In October 2010, \LIGO{} completed its sixth science run
(S6) and Virgo completed its third science run (VSR3).  While both
\LIGO{} detectors and Virgo were operating, several all-sky detection
pipelines operated in a low-latency configuration to send astronomical alerts,
namely \textsc{mbta}, Coherent WaveBurst, and
Omega~\cite{HugheyGWPAW2011, S6lowlatency}.
\editorial{Get references for these low-latency pipelines.} The S6 analyses
achieved latencies of 30--60 minutes, which were dominated by a human vetting
process. Candidates were sent for \EM{} followup to several
telescopes; Swift, \textsc{rotse}, \textsc{tarot}, and Zadko~\cite{kanner2008,
HugheyGWPAW2011} took images of likely sky locations.  \textsc{mbta} achieved
the best \GW{} trigger-generation latencies of 2--5 minutes.  We
assume that in the advanced detector era the vetting process will be automated,
so current \GW{} search methodology and telescope actuation would
dominate latency.

Realizing advance detection of compact binary coalescences (\CBC{}s) will
require striking a balance between latency and throughput. \CBC{} searches
consist of banks of matched filters, or
cross-correlations between the data stream and a bank of nominal ``template''
signals.  There are many different implementations of matched filters, but most
have high throughput at the cost of high latency, or low latency at the cost of
low throughput.  The former are epitomized by the overlap-save algorithm
\cite{numerical-recipes-chapter-13} for frequency domain (\FD) convolution,
currently the preferred method in \GW{}
searches.  The most obvious example of the latter is the time domain
(\TD) convolution, which has no algorithmic latency.  However, its
computational complexity is quadratic in the length of the templates, so it is
prohibitively expensive for long templates.

Fortunately, the morphology of inspiral signals can be exploited to offset some
of the computational complexity of low-latency algorithms.  First, the signals
evolve slowly in frequency, so that they can be broken into contiguous
band-limited time intervals and processed at possibly lower sample rates.
Second, inspiral filter banks consist of highly similar templates, admitting
methods such as singular value decomposition (\SVD{}) to reduce the number of
templates \cite{Cannon:2010p10398}. We will use both aspects to demonstrate that a very
low latency analysis with advance detection of compact binary sources is
possible with current computing resources.  Assuming other technical sources of
latency can be reduced significantly, this should allow the possibility for
prompt alerts to be sent to the astronomical community.

>>>>

The coalescence of compact binary systems consisting of neutron stars
 (\textsc{ns}) and/or black holes (\textsc{bh}) is presently the best
understood and most actively pursued target source of gravitational
radiation for ground-based gravitational-wave (\GW{}) detectors.
As a compact binary system loses energy to \GW{}s, its
orbital separation decays, leading to a run-away inspiral with the \GW{}
amplitude and frequency increasing until the system eventually merges.  It is
thought that if a \textsc{ns} is involved, it may become tidally disrupted near
the merger and fuel a bright electromagnetic (\EM{})
counterpart~\citep{shibata:2007}. Observation of an \EM{} counterpart to a
candidate \GW{} event will vastly boost confidence in the \GW{} detection and
will often provide sufficient sky localization to determine
the source's host galaxy and therefore measure the source's redshift. The
\GW{} observation of this event will also provide a luminosity
distance measurement. With both quantities, we can
immediately measure the Hubble constant with unprecedented
accuracy~\citep{2010ApJ...725..496N}. The most efficient means of obtaining
such joint observations is through the rapid identification and dissemination
of candidate \GW{} events.

In the era of first-generation detectors, the \GW{} community initiated a
project to send alerts when potential \GW{} transients were observed in
order to trigger followup observations by optical telescopes.  The best
demonstrated latencies were on the order of 1 hour%
~\citep{HugheyGWPAW2011}, which was an important
achievement, but far too late to catch prompt \EM{}
emission. In order to extract the maximum science from Advanced \LIGO{}
\CBC{} events, we have the ambition of reporting \GW{} candidates not minutes
\emph{after} the merger, as has already been accomplished, but seconds
\emph{before}.

Though compact binary coalescence (\CBC{}) events may be responsible for several
classes of \EM{} transients, \CBC{}s are in particular believed to be a
mechanism for short gamma-ray bursts (short \GRB{}s)~\citep{Lee:2005, nakar07}.
In this scenario, prompt \EM{} emission arises as shells of
relativistically out-flowing matter collide in the inner shock.
The same inner shocks, or potentially reverse shocks, can produce a bright
accompanying optical flash~\citep{Sari99}. Prompt emission is a probe into
the extreme initial conditions of the outflow, in contrast with afterglows,
which arise in the external shock with the local medium and are relatively
insensitive to initial conditions.

Optical flashes have only been observed for a handful of long
\GRB{}s~\citep{2011CRPhy..12..255A} by telescopes with extremely rapid response
or, in the case of \textsc{grb 080319b}, by pure serendipity, where several
telescopes were observing a previous \GRB{} in the same
field~\citep{2008Natur.455..183R}. The observed optical flashes peaked
within tens of seconds and decayed quickly.
Short \GRB{}s, on the other hand, typically
fade too quickly to even catch the tails of the afterglows in any band. Rapid
\GW{} transient alerts could enable the first observation of
prompt optical flashes and the rise of afterglows from short \GRB{}s, and, for
an exceptional event, perhaps even a glimpse of a
precursor~\citep{0004-637X-723-2-1711}.

We show that by examining the signal to noise ratio (\SNR{}) stream for
threshold crossings before the \GW{} signal leaves the detection band, it is
possible to trade some \SNR{} and sky localization
accuracy for negative latency.  Figure~\ref{fig:earlywarning} shows
projected early trigger rates for \textsc{ns}--\textsc{ns} binaries in Advanced
\LIGO\ assuming event rates predicted in~\citep{Abadie:2010p10836} and anticipated
detector sensitivity described in \citep{ALIGONoise}.  The
most likely estimates indicate that $\sim$10 sources will produce an \SNR\
above 8 in the detector 10 seconds or more prior to the merger during one
year of observation, though the uncertainty spans orders of magnitude.
The gray bands give the 5 to 95\% confidence interval as described
in~\citep{Abadie:2010p10836}.
%
\begin{figure}[h]
\plotone{figures/snr_in_time.pdf}
\caption{\label{fig:earlywarning} Expected number of \textsc{ns}--\textsc{ns}
sources that will be detectable $t$ seconds before coalescence.  The heavy
solid line is the most likely yearly rate estimate $\dot N_{\mathrm{re}}$
during Advanced \LIGO.  The shaded region represents the confidence interval
$[\dot N_{\mathrm{low}}, \dot N_{\mathrm{high}}]$ described in
\citep{Abadie:2010p10836}.  With an advanced detector in the `zero detuning,
high power' configuration \citep{ALIGONoise} and an \SNR\ threshold of 8,
we will observe a total of $\dot N_{\mathrm{re}} = 40$ events~yr$^{-1}$, but
$\sim$10~yr$^{-1}$ will be detectable within 10~s of merger and
$\sim$1~yr$^{-1}$ within 100~s.}
\end{figure}

In October 2010, \LIGO{} completed its sixth science run
(S6) and Virgo completed its third science run (VSR3).  While both
\LIGO{} detectors and Virgo were operating, several all-sky detection
pipelines operated in a low-latency configuration to send astronomical alerts,
namely \textsc{mbta}, Coherent WaveBurst, and
Omega~\citep{HugheyGWPAW2011, S6lowlatency}. The S6 analyses
achieved latencies of 30--60 minutes, which were dominated by a human vetting
process. Candidates were sent for \EM{} followup to several
telescopes; Swift, \textsc{rotse}, \textsc{tarot}, and Zadko~\citep{kanner2008,
HugheyGWPAW2011} took images of likely sky locations.  \textsc{mbta} achieved
the best \GW{} trigger-generation latencies of 2--5 minutes.  We
assume that in the advanced detector era the vetting process will be automated,
so current \GW{} search methodology and telescope actuation would
dominate latency.

Realizing advance detection of compact binary coalescences (\CBC{}s) will
require striking a balance between latency and throughput. \CBC{} searches
consist of banks of matched filters, or
cross-correlations between the data stream and a bank of nominal ``template''
signals.  There are many different implementations of matched filters, but most
have high throughput at the cost of high latency, or low latency at the cost of
low throughput.  The former are epitomized by the overlap-save algorithm
\citep{numerical-recipes-chapter-13} for frequency domain (\FD) convolution,
currently the preferred method in \GW{}
searches.  The most obvious example of the latter is the time domain
(\TD) convolution, which has no algorithmic latency.  However, its
computational complexity is quadratic in the length of the templates, so it is
prohibitively expensive for long templates.

Fortunately, the morphology of inspiral signals can be exploited to offset some
of the computational complexity of low-latency algorithms.  First, the signals
evolve slowly in frequency, so that they can be broken into contiguous
band-limited time intervals and processed at possibly lower sample rates.
Second, inspiral filter banks consist of highly similar templates, admitting
methods such as singular value decomposition (\SVD{}) to reduce the number of
templates \citep{Cannon:2010p10398}. We will use both aspects to demonstrate that a very
low latency analysis with advance detection of compact binary sources is
possible with current computing resources.  Assuming other technical sources of
latency can be reduced significantly, this should allow the possibility for
prompt alerts to be sent to the astronomical community.
